---
title: "Fox Discussion Week1"
date: 1/29/2025
format: docx
---

Hello all – I chose a three-part vector exercise this week for my first attempt at using R for linear algebra problems. I explored a few packages and struggled with unexpected results when finding the null set in part 1, but eventually found a solution and the rest went quite well. 

###Problem: 

```{r libraries, message = FALSE, warning = FALSE}
# load libraries
library(tidyverse)
library(MASS)
library(pracma)
```

First I created the matrix A and vector z:  
```{r matrix}
# create matrix A
A <- matrix(c(2,3,1,4,
              1,2,1,3,
              -1,0,1,1), 
            nrow = 3,
            byrow = TRUE)
A

# create vector z
z <- c(3,-5,1,2)
z
```
#### 1. Find a set S so that N(A) = <S> 

This was unusually challenging. First I used the MASS package: the Null() function should find the null space of A in a single step, outputting the set of basis vectors. However, it provided only one vector, in decimals:

```{r MASS}
A_null_MASS <- Null(A) 
A_null_MASS
```
This single vector is not the correct solution as there are two free variables shown in the row reduced matrix below (using the pracma package):

```{r RREF}
A_rref <- rref(A)
A_rref
```
Manually, I got this set of two basis vectors and verified with an online calculator, which gave me a way to check my results from R: 
```{r manual_solution}
manual_solution <- matrix(c(1,-1,1,0,
                            1,-2,0,1),
                          ncol=2)
manual_solution
```
The pracma package did a little better than MASS as it did result in two vectors, however it was still displayed as decimals. I believe this has to do with scaling but I was not able to reverse to get the expected integer values:
```{r pracma}
A_null_pracma <- nullspace(A)
A_null_pracma
```
Finally I found a solution on stackoverflow.com (Stack Overflow). A kind soul had created the below formula which easily returned the expected basis vectors. It uses the QR decomposition method which is said to be more stable. I am still digesting it, but I appreciate that it is transparent and easily provided the expected response, which will help my learning.

```{r kindsoul}
# "https://stackoverflow.com/questions/43223579/solve-homogenous-system-ax-0-for-any-m-n-matrix-a-in-r-find-null-space-basi)"

NullSpace <- function (A) {
  m <- dim(A)[1]; n <- dim(A)[2]
  ## QR factorization and rank detection
  QR <- base::qr.default(A)
  r <- QR$rank
  ## cases 2 to 4
  if ((r < min(m, n)) || (m < n)) {
    R <- QR$qr[1:r, , drop = FALSE]
    P <- QR$pivot
    F <- R[, (r + 1):n, drop = FALSE]
    I <- base::diag(1, n - r)
    B <- -1.0 * base::backsolve(R, F, r)
    Y <- base::rbind(B, I)
    X <- Y[base::order(P), , drop = FALSE]
    return(X)
  }
  ## case 1
  return(base::matrix(0, n, 1))
}

#Call function
S <- NullSpace(A)
round(S, 15)

S
```
### 2. Determine if vector z is in spanning set S

By multiplying A by z, the zero vector is returned, which means that z is in the null space of A (spanning set S):
```{r zero_vector}
# Multiply A by z
test_z <- A %*% z
test_z
```

### 3. Write as a linear combination

Since z is in the null space of A (spanning set S), we can find the coefficients for the two basis vectors in <S> that will give us vector z. R provides the "qr.solve" function, which returns the below coefficients, giving us the linear combination:

z =  S1 +  2(S2)

```{r coefficients}
c <- qr.solve(S,z)
c
```

### 4. Practical Applications

While this was just an abstract exercise, null space, spanning sets, and linear combinations are foundational concepts. I have encountered these concepts in system optimization problems for example, and also in identifying redundant variables in predictive modeling. I understand they are also found in engineering problems, image compression and noise filtering (redundant information again). Knowing the math is helpful in better understanding these and other applications. 
Regarding challenges in real-world applications, I would venture to say data quality and availability, computing power, and practicality of the solution would all come into play.